Type theories as quotient inductive-recursive types

### 2025-01-23 tex/main.tex
Fix the coherence statement in S3.4

One possibility is to define substitution calculus as a quotient inductive-inductive-\emph{recursive} type whereas type substitution as a simultaneously defined function computes.
Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ along do not make $[\blank]_{\Ty}\blank$ a function, we may define it by
\begin{alignat*}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
[ \sigma ]_{\Ty}\;\UU & = \UU
\end{alignat*}
However, this definition is \emph{not} inductive-recursive in the sense of Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}: in their schema, the codomain of a simultaneously defined function does not refer to any inductive type being defined.
While its general schema remains underdeveloped, this definition is accepted in proof assistants such as \Agda and used in other formalisation of type theory~\cite{Danielsson2006}.
Still, although $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ can be proved equal to $A$ and $[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, they are not definitional but propositional.
This definition fails to eliminate any transport in the definition of substitution calculus.

Our second attempt is to define type substitution with additional rules:
\begin{alignat}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
[ \sigma ]_{\Ty}       \;\UU  & = \UU                                   \label{eq:type-sub-at2-1} \\
[ \idS ]_{\Ty}         \;A    & = A                                     \label{eq:type-sub-at2-2} \\
[ \sigma ; \tau ]_{\Ty}\;A    & = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
\end{alignat}
However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} definitionally in a type-theoretic proof assistant.
By design, definitions by pattern matching will be translated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} or an explicit construct for pattern matching, so only the eliminator for $\Ty\;\Delta$ will be considered in the above definition.
\LT[noinline]{How about \Coq and \Lean?}
Using the first-match semantics, the above definition amounts to the following definition:
\begin{alignat*}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
\end{alignat*}
where $\reduce$ is used to emphasise the reduction (which entails the definitional equality): we have done nothing different from our first attempt! 
%As a result, neither  are definitional.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} definitional, but then \eqref{eq:type-sub-at2-1} will be translated to 
\begin{alignat*}{3}
[ \emptysub ]_{\Ty}\;\UU         & \reduce \UU \\
[ \sigma, t ]_{\Ty}\;\UU         & \reduce \UU \\
[ \pi_1\,\sigma ]_{\Ty}\;\UU     & \reduce \UU,
\end{alignat*}
making $[\sigma]\;\UU = \UU$ propositional instead.
Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
That is, using the standard semantics of function definition, we are not able to make type substitution definitional on \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$.

To ensure that type substitution is definitional for all \emph{rules}, we turn to a definition by overlapping patterns (\Cref{sec:meta:overlapping}) instead to make each clause definitional. 
In addition, we can include other derived rules such as $[\pi_1(\sigma, t)]_{\Ty}\;A = [ \sigma]_{\Ty}\;A$ as long as it is locally confluent and terminating.
We then end up with the following definition:
\begin{alignat}{3}
[\blank]_{\Ty} \blank            &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \nonumber \\
[ \idS ]_{\Ty}\;A                & \reduce A \label{eq:def-type-subst-1}\\
[ \sigma ; \tau ]_{\Ty}\;A       & \reduce [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:def-type-subst-2}\\
[ \pi_1(\sigma, t) ]_{\Ty}\;A    & \reduce [\sigma]_\Ty\;A                       \label{eq:def-type-subst-3} \\
[ \pi_1(\sigma; \tau) ]_{\Ty}\;A & \reduce [\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)    \label{eq:def-type-subst-4} \\
[ \sigma ]_{\Ty}\;\UU            & \reduce \UU                                   \label{eq:def-type-subst-5}
\end{alignat}
We first check that this definition of type substitution is confluent.\footnote{%
\Agda implements a local confluence checker with the experimental option \text{-{}-local-confluence-check} which is turned on globally in our formalisation, but we do not trust experimental features blindly.}
\begin{proposition}[Local confluence] \label{prop:local-confluence-1}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  To show the local confluence, we consider all local peaks, i.e.\ each combination of \eqref{eq:def-type-subst-5} with \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4},
  \[
    [\idS]_{\Ty}\;\UU, \qquad [\sigma;\tau]_{\Ty}\;\UU, \qquad [\pi_1(\sigma, t)]_{\Ty}\;\UU, \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;\UU
  \]
  and show that each applicable clause does result in $\UU$.

@@

Yet, as we are defining a function on an inductive type with equality constructors, we have to prove that $[\blank]_{\Ty}\blank$ is coherent.\LT{general condition?} 
This boils down to that $[\blank]_{\Ty}\blank$ preserves equality constructors for quotient types.
\begin{proposition}[Coherence]\label{prop:coherence-1}
  $[\sigma]_{\Ty}\,A$ is (propositionally) equal to $[\tau]_{\Ty}\,A$ whenever $\sigma = \tau$.
  \danger
\end{proposition}
\begin{proof}
  As equality constructors $[\idS]_T$, $[;]_T$, and $[]\UU$ for types become definitional, it suffices to show that following identities
  \begin{align*}
    [\sigma;\idS]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [\idS;\sigma]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [(\sigma;\tau);\gamma]_{\Ty} \;A & = [\sigma; (\tau; \gamma)]_{\Ty}A, \\
    [\pi_1(\sigma, t)]_{\Ty}\;A      & = [\sigma]_{\Ty}\;A, \\
    [\sigma; (\tau, t)]_{\Ty} \;A    & = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\emptyctx]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
  \end{align*}
  hold (propositionally).
  The first two identities of the first row hold definitionally by \eqref{eq:def-type-subst-1} and \eqref{eq:def-type-subst-2}, while terms of the both sides of the third reduce to $[\sigma]_{\Ty}\;([\tau]_{\Ty}\;([\gamma]_{\Ty}\;A))$ by \eqref{eq:def-type-subst-2}.

@@

In this section, we extend substitution calculus with an empty universe $\UU$ of small types~$\El\,u$.
For a QII definition \cite{Altenkirch2016a}, this extension amounts to adding the following constructors 
\begin{alignat*}{3}
  \UU   & : \implicit{\Gamma} && \Ty\, \Gamma \\
  \El   & : \implicit{\Gamma} && \Tm\,\Gamma\;\UU \to \Ty\,\Gamma \\
  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && [ \sigma ]\,(\El\, u) =^{\Tm\;\Gamma\;\UU} \El\,(\transfib{\Tm\;\Gamma}{[]\UU}{[\sigma]{u}})
\end{alignat*}
where the type substitution $[\sigma](\El\,u)$ is, informally, the type of the term substitution $[\sigma]\,u$ for $u : \Tm\;\Delta\;\UU$, and the transport is needed to coerce $[\sigma]\,u : \Tm\;\Gamma\;([\sigma]\;\UU)$ into $\Tm\;\Gamma\;\UU$.

In our QIIR definition, the transported term $\transfib{\Tm\;\Gamma}{[]\UU}{[\sigma]{u}}$ is no longer needed, since the equality constructor $[]\UU$ becomes definitional. 
Therefore, we may replace the equality constructor $[]\El$ by a clause of type substitution:
\begin{alignat}{3}
  [ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma] {u}) \label{eq:def-type-subst-6}
\end{alignat}
Yet, this naive change breaks the local confluence!
Observe that the local peak
\[
  \El\,u \Leftarrow [\idS]_{\Ty}(\El\,u) \reduce \El([\idS]\,u)
\]
cannot reduce to the same term, since the term substitution $[\idS]\,u$ is merely a constructor.

@@

Then, \eqref{eq:def-type-subst-6} and the rule $\mathsf{concat}$ are accordingly changed to 
\begin{align}
  [ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma]_{\Tm} {u}) \label{eq:def-type-subst-7}\\
  \mathsf{concat} : \sigma ; (\tau , t) & =^{\Sub\,\Gamma\,(\Theta, A)} (\sigma ; \tau) , [ \sigma ]_{\Tm} t \nonumber
\end{align}
to retain the local confluence for type substitution:
\begin{proposition}[Local confluence] \label{prop:local-confluence-2}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  In addition to previous cases in the proof of \Cref{prop:local-confluence-1}, consider following terms 
  \[
    [\idS]_{\Ty}\;(\El\,u), \qquad [\sigma;\tau]_{\Ty}\;(\El\,u), \qquad [\pi_1(\sigma, t)]_{\Ty}\;(\El\,u), \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;(\El\,u)
  \]
  which reduce to the following terms
  \[
    \El\,u, \qquad \El([\sigma]_{\Tm}\,[\tau]_{\Tm}\,u), \qquad \El([\sigma]_{\Tm}\,u), \qquad\text{and}\qquad
    \El([\sigma]_{\Tm}\,[\pi_1\tau]_{\Tm}\,u)
  \]
  respectively, no matter which function clause of $[\blank]_{\Ty}\blank$ is used.
  Therefore, type substitution is locally confluent.
\end{proof}
\begin{proposition}[Termination]
  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
  \danger
\end{proposition}
\begin{proof}
  \LT{clearly terms on the RHS are structurally smaller~\cite{Abel2002}.}
\end{proof}

@@

\begin{proposition}\label{prop:correctness-1}
  For any $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Tm\;\Delta\;A$, the term $[\sigma]t$ is propositionally equal to $[\sigma]_{\Tm}\,t$.
  \danger
\end{proposition}
\begin{proof}
  We show this statement by induction on $\sigma$.
  \begin{enumerate}
    \item For the identity substitution $\idS$, we have $[\idS]t : [\idS]t = t$ but $t \judgeq [\idS]_{\Tm} t$ by definition.
    \item For a substitution composite $\sigma; \tau$, we have
      \begin{align*}
        [\sigma;\tau]t & = [\sigma]\,[\tau] t               && \text{by $[;]t$} \\
                       & = [\sigma]\,([\tau]_{\Tm}\,t)       && \text{by the induction hypothesis $[\tau]t = [\tau]_{\Tm}\,t$ for any $t$} \\
                       & = [\sigma]_{\Tm}\,([\tau]_{\Tm}\,t) && \text{by the induction hypothesis $[\sigma]u = [\sigma]_{\Tm}\,u$ for any $u$} \\
                       & \judgeq [\sigma;\tau]_{\Tm}\,t && \text{by definition.}
      \end{align*}
    \item For $\pi_1(\sigma, t)$, we have
      \begin{align*}
        [\pi_1(\sigma, t)]t & = [\sigma] t               && \text{by $\pi_1\beta$} \\
                            & = [\sigma]_{\Tm} \,t       && \text{by the induction hypothesis $[\tau]t = [\tau]_{\Tm}\,t$ for any $t$} \\
                            & \judgeq [\pi_1(\sigma, t)]_{\Tm}\,t && \text{by definition.}
      \end{align*}
    \item For $\pi_1(\sigma; \tau)$, it is proved similarly as the previous case.
    \item For the remaining cases, $[\sigma]_{\Tm} t$ is equal to $[\sigma]\,t$ definitionally.
  \end{enumerate}
  Hence, we conclude that $[\blank]_{\Tm}\blank$ is propositionally equal to $[\blank]\blank$.
  \LT[noinline]{Shall we say anything about equality constructors?}
\end{proof}

Finally, we have to check that both $[\blank]_{\Ty}\blank$ and $[\blank]_{\Tm}\blank$ are coherent with respect to equality constructors.
\begin{proposition}[Coherence] \label{prop:coherence-2}
  The following statements hold:
  \danger
  \begin{enumerate}
    \item $[\sigma]_{\Ty}\,A = [\tau]_{\Ty}\,A$ whenever $\sigma = \tau$.
    \item $[\sigma]_{\Tm}\,A = [\tau]_{\Tm}\,A$ whenever $\sigma = \tau$.
  \end{enumerate}
\end{proposition}
\begin{proof}
  By \Cref{re:coherence-proof}, for type substitution, we only have to consider the following cases 
  \begin{align*}
    [\sigma; (\tau, t)]_{\Ty} \;A    & = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\emptyctx]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
  \end{align*}
  while other cases hold definitionally.
  Again, for each case, we prove them by induction on $A$ and consider the case $A \defeq \El\,u$ only which is added to the definition of $[\blank]_{\Ty}\blank$ in this section. 
  However, all cases follow easily from \Cref{prop:correctness-1}.
  For example, consider the first case:
  \begin{align*}
    [\sigma; (\tau, t)]_{\Ty}(\El\,u) & \judgeq \El([\sigma; (\tau, t)]_{\Tm}\,u) && \text{by definition \eqref{eq:def-type-subst-7}} \\
                                      & = \El( [\sigma; (\tau, t)]\,u)            && \text{by \cref{prop:correctness-1}} \\
                                      & = \El( [\sigma; \tau, [\sigma]t)]\,u)     && \text{by $\mathsf{concat}$} \\
                                      & = \El( [\sigma; \tau, [\sigma]t)]_{\Tm}\,u) && \text{by \cref{prop:correctness-1}} \\
                                      & \judgeq [\sigma; \tau, [\sigma]t)]_{\Ty}\left(\El\,u\right) && \text{by definition \eqref{eq:def-type-subst-7} }
  \end{align*}
  The remaining two cases are omitted.

@@

We have seen that the extension of substitution calculus with an (empty) universe breaks the local confluence of type substitution defined in \cref{subsec:SC-QIIRT}, because the newly added function clause interacts with other existing clauses.
In general, every term occurs on the right-hand side of a clause needs to propagate the reduction to allow a confluent term to exist, and this desired propagation can be achieved by turning a term involving a substitution $\sigma$ into a recursion.
We will see the same pattern in the next section.

@@

\begin{proposition} \label{prop:correctness-2}
  For every $\sigma : \Sub\;\Gamma\;\Delta$, and $A : \Ty\;\Delta$
  \danger
  $\sigma \uparrow A$ is propositionally equal to $(\pi_1 \idS ; \sigma , \pi_2 \idS)$.
\end{proposition}
\begin{proof}
  We prove the statement by induction on $\sigma$ with \cref{lem:lifting}.
  Every case follows from equational reasoning straightforwardly.
  For example, if $\sigma = \sigma;\tau$ is a composite, we have
  \begin{align*}
    (\sigma;\tau) \uparrow A & \judgeq (\sigma \uparrow [\tau]\,A);(\tau \uparrow A) && \text{by definition} \\
                             & = \sigma^+;\tau^+                                     && \text{by induction hypothesis} \\ 
                             & = (\sigma;\tau)^+                                     && \text{by \cref{lem:lifting}.} 
  \end{align*}
  Other cases follow similarly.
\end{proof}

>>>>

One possibility is to define substitution calculus as a QII\emph{R} type where type substitution, as a simultaneously defined recursion, does compute.
Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may define it by
\begin{alignat*}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
[ \sigma ]_{\Ty}\;\UU & = \UU
\end{alignat*}
Note that this definition is \emph{not} inductive-recursive in the sense of Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}: in their schema, the codomain of a simultaneously defined function does not refer to any inductive type being defined.\footnote{%
  While the general schema for this more liberal notion of induction-recursion remains underdeveloped, this definition is accepted in \Agda and used in other formalisation of type theory~\cite{Danielsson2006}.}
Still, although $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ can be proved equal to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, they are not definitional but propositional.
This definition fails to eliminate any transport in the definition of substitution calculus.

Our second attempt is to define type substitution with additional rules:
\begin{alignat}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
[ \sigma ]_{\Ty}       \;\UU  & = \UU                                   \label{eq:type-sub-at2-1} \\
[ \idS ]_{\Ty}         \;A    & = A                                     \label{eq:type-sub-at2-2} \\
[ \sigma ; \tau ]_{\Ty}\;A    & = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
\end{alignat}
However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} definitionally.
By design, definitions by pattern matching will be translated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} or an explicit construct for pattern matching, so only the eliminator for $\Ty\;\Delta$ will be considered in the above definition.
\LT[noinline]{How about \Coq and \Lean?}
Using the first-match semantics, the above definition amounts to the following definition:
\begin{alignat*}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
\end{alignat*}
where $\reduce$ is used to emphasise the reduction (which entails the definitional equality): we have done nothing different from our first attempt! 
%As a result, neither  are definitional.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} definitional, but then \eqref{eq:type-sub-at2-1} will be translated to 
\begin{alignat*}{3}
[ \emptysub ]_{\Ty}\;\UU         & \reduce \UU \\
[ \sigma, t ]_{\Ty}\;\UU         & \reduce \UU \\
[ \pi_1\,\sigma ]_{\Ty}\;\UU     & \reduce \UU,
\end{alignat*}
making $[\sigma]\;\UU = \UU$ propositional instead.
Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
That is, using the standard semantics of function definition, we are not able to make type substitution definitional on \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$.

To ensure that type substitution is definitional for all \emph{rules}, we turn to a definition by overlapping patterns (\Cref{sec:meta:overlapping}) instead to make each clause definitional. 
In addition, we can include other derived rules such as $[\pi_1(\sigma, t)]_{\Ty}\;A = [ \sigma]_{\Ty}\;A$ as long as it is locally confluent and terminating.
We end up with the following definition for type substitution:
\begin{alignat}{3}
[\blank]_{\Ty} \blank            &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \nonumber \\
[ \idS ]_{\Ty}\;A                & \reduce A \label{eq:def-type-subst-1}\\
[ \sigma ; \tau ]_{\Ty}\;A       & \reduce [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:def-type-subst-2}\\
[ \pi_1(\sigma, t) ]_{\Ty}\;A    & \reduce [\sigma]_\Ty\;A                       \label{eq:def-type-subst-3} \\
[ \pi_1(\sigma; \tau) ]_{\Ty}\;A & \reduce [\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)    \label{eq:def-type-subst-4} \\
[ \sigma ]_{\Ty}\;\UU            & \reduce \UU                                   \label{eq:def-type-subst-5}
\end{alignat}
We first check that this definition of type substitution is confluent.\footnote{%
\Agda implements a local confluence checker with the experimental option \text{-{}-local-confluence-check} which is turned on globally in our formalisation, but we do not trust experimental features blindly.}
\begin{proposition}[Local confluence] \label{prop:local-confluence-1}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  To show the local confluence, we consider all local peaks, i.e.\ each combination of \eqref{eq:def-type-subst-5} with \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4},
  \[
    [\idS]_{\Ty}\;\UU, \qquad [\sigma;\tau]_{\Ty}\;\UU, \qquad [\pi_1(\sigma, t)]_{\Ty}\;\UU, \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;\UU
  \]
  and show that each applicable clause does result in $\UU$.

@@

Yet, as we are defining a function on a quotient inductive type, we have to prove that $[\blank]_{\Ty}\blank$ is coherent.
This boils down to that $[\blank]_{\Ty}\blank$ preserves all equality constructors.
\LT{Should we mention general condition for coherence?} 
\begin{proposition}[Coherence]\label{prop:coherence-1}
  $[\sigma]_{\Ty}\,A$ is (propositionally) equal to $[\tau]_{\Ty}\,A$ for every equality constructor $p : \sigma = \tau$.
  \danger
\end{proposition}
\begin{proof}
  As equality constructors $[\idS]_T$, $[;]_T$, and $[]\UU$ for types become definitional, it suffices to show that following identities
  \begin{align*}
    [\sigma;\idS]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [\idS;\sigma]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [(\sigma;\tau);\gamma]_{\Ty} \;A & = [\sigma; (\tau; \gamma)]_{\Ty}A, \\
    [\pi_1(\sigma, t)]_{\Ty}\;A      & = [\sigma]_{\Ty}\;A, \\
    [\sigma; (\tau, t)]_{\Ty} \;A    & = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\emptyctx]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
  \end{align*}
  hold (propositionally).
  The first two identities of the first row hold definitionally by \eqref{eq:def-type-subst-1} and \eqref{eq:def-type-subst-2}, while terms of the both sides of the third reduce to $[\sigma]_{\Ty}\;([\tau]_{\Ty}\;([\gamma]_{\Ty}\;A))$ by \eqref{eq:def-type-subst-2}.

@@

In this section, we extend substitution calculus with an empty universe $\UU$ of small types~$\El\,u$.
For a QII definition \cite{Altenkirch2016a}, this extension amounts to adding the following constructors 
\begin{alignat*}{3}
  \UU   & : \implicit{\Gamma} && \Ty\, \Gamma \\
  \El   & : \implicit{\Gamma} && \Tm\,\Gamma\;\UU \to \Ty\,\Gamma \\
  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && [ \sigma ]\,(\El\, u) =^{\Tm\;\Gamma\;\UU} \El\,(\alert{\transfib{\Tm\;\Gamma}{[]\UU}{\color{black}[\sigma]{u}}})
\end{alignat*}
where the type substitution $[\sigma](\El\,u)$ is, informally, the type of the term substitution $[\sigma]\,u$ for $u : \Tm\;\Delta\;\UU$, and the transport is needed to coerce $[\sigma]\,u : \Tm\;\Gamma\;([\sigma]\;\UU)$ into $\Tm\;\Gamma\;\UU$.

In our QIIR definition, $\transfib{\Tm\;\Gamma}{[]\UU}{\blank}$ is no longer needed, since the equality constructor $[]\UU$ becomes definitional. 
Therefore, the equality constructor $[]\El$ may be replaced by a clause of type substitution:
\begin{alignat}{3}
  [ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma] {u}) \label{eq:def-type-subst-6}
\end{alignat}
Yet, this naive change breaks the local confluence!
Observe that the local peak
\[
  \El\,u \Leftarrow [\idS]_{\Ty}(\El\,u) \reduce \El([\idS]\,u)
\]
cannot reduce to the same term, since the term substitution $[\idS]\,u$ is merely a constructor.

@@

Then, \eqref{eq:def-type-subst-6} and the rule $\mathsf{concat}$ are accordingly changed to 
\begin{align}
  [ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma]_{\Tm} {u}) \label{eq:def-type-subst-7}\\
  \mathsf{concat} : \sigma ; (\tau , t) & =^{\Sub\,\Gamma\,(\Theta, A)} (\sigma ; \tau) , [ \sigma ]_{\Tm} t \nonumber
\end{align}
to retain the local confluence for type substitution:
\begin{proposition}[Local confluence] \label{prop:local-confluence-2}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  In addition to previous cases in the proof of \Cref{prop:local-confluence-1}, consider terms 
  \[
    [\idS]_{\Ty}\;(\El\,u), \qquad [\sigma;\tau]_{\Ty}\;(\El\,u), \qquad [\pi_1(\sigma, t)]_{\Ty}\;(\El\,u), \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;(\El\,u)
  \]
  which reduce to the following terms
  \[
    \El\,u, \qquad \El([\sigma]_{\Tm}\,[\tau]_{\Tm}\,u), \qquad \El([\sigma]_{\Tm}\,u), \qquad\text{and}\qquad
    \El([\sigma]_{\Tm}\,[\pi_1\tau]_{\Tm}\,u)
  \]
  respectively, no matter which function clause of $[\blank]_{\Ty}\blank$ is used.
  Therefore, type substitution is locally confluent.
\end{proof}
\begin{proposition}[Termination]
  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
  \danger
\end{proposition}
\begin{proof}
  \LT{clearly terms on the RHS are structurally smaller~\cite{Abel2002}.}
\end{proof}

@@

\begin{proposition}\label{prop:correctness-1}
  For any $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Tm\;\Delta\;A$, the term $[\sigma]t$ is propositionally equal to $[\sigma]_{\Tm}\,t$.
\end{proposition}
\begin{proof}
  We show this statement by induction on $\sigma$.
  \begin{enumerate}
    \item For the identity substitution $\idS$, we have $[\idS]t : [\idS]t = t$ but $t \judgeq [\idS]_{\Tm} t$ by definition.
    \item For a substitution composite $\sigma; \tau$, we have
      \begin{align*}
        [\sigma;\tau]t & = [\sigma]\,[\tau] t               && \text{by $[;]t$} \\
                       & = [\sigma]\,([\tau]_{\Tm}\,t)       && \text{by the induction hypothesis $[\tau]t = [\tau]_{\Tm}\,t$ for any $t$} \\
                       & = [\sigma]_{\Tm}\,([\tau]_{\Tm}\,t) && \text{by the induction hypothesis $[\sigma]u = [\sigma]_{\Tm}\,u$ for any $u$} \\
                       & \judgeq [\sigma;\tau]_{\Tm}\,t && \text{by definition.}
      \end{align*}
    \item For $\pi_1(\sigma, t)$, we have
      \begin{align*}
        [\pi_1(\sigma, t)]t & = [\sigma] t               && \text{by $\pi_1\beta$} \\
                            & = [\sigma]_{\Tm} \,t       && \text{by the induction hypothesis $[\tau]t = [\tau]_{\Tm}\,t$ for any $t$} \\
                            & \judgeq [\pi_1(\sigma, t)]_{\Tm}\,t && \text{by definition.}
      \end{align*}
    \item For $\pi_1(\sigma; \tau)$, it is proved similarly as the previous case.
    \item For the remaining cases, $[\sigma]_{\Tm} t$ is equal to $[\sigma]\,t$ definitionally.
  \end{enumerate}
  Hence, we conclude that $[\blank]_{\Tm}\blank$ is propositionally equal to $[\blank]\blank$.
  \LT[noinline]{Shall we say anything about equality constructors?}
\end{proof}

Finally, we have to check that both $[\blank]_{\Ty}\blank$ and $[\blank]_{\Tm}\blank$ are coherent with respect to equality constructors.
\begin{proposition}[Coherence] \label{prop:coherence-2}
  For every equality constructor $p : \sigma = \tau$, the following identities
  \danger
  \[
    [\sigma]_{\Ty}\,A = [\tau]_{\Ty}\,A
    \quad\text{and}\quad
    [\sigma]_{\Tm}\,t = [\tau]_{\Tm}\,t
  \]
  for any type $A$ and any term $t$.
\end{proposition}
\begin{proof}
  By \Cref{re:coherence-proof}, for type substitution, we only have to consider the following cases 
  \begin{align*}
    [\sigma; (\tau, t)]_{\Ty} \;A    & = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\emptyctx]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
  \end{align*}
  while other cases hold definitionally.
  Again, for each case, we prove them by induction on $A$ and consider the case $A \defeq \El\,u$ only which is added to the definition of $[\blank]_{\Ty}\blank$ in this section. 
  However, all cases follow easily from \Cref{prop:correctness-1}.
  For example, consider the first case:
  \begin{align*}
    [\sigma; (\tau, t)]_{\Ty}(\El\,u) & \judgeq \El([\sigma; (\tau, t)]_{\Tm}\,u) && \text{by definition \eqref{eq:def-type-subst-7}} \\
                                      & = \El( [\sigma; (\tau, t)]\,u)            && \text{by \cref{prop:correctness-1}} \\
                                      & = \El( [\sigma; \tau, [\sigma]t)]\,u)     && \text{by $\mathsf{concat}$} \\
                                      & = \El( [\sigma; \tau, [\sigma]t)]_{\Tm}\,u) && \text{by \cref{prop:correctness-1}} \\
                                      & \judgeq [\sigma; \tau, [\sigma]t)]_{\Ty}\left(\El\,u\right) && \text{by definition \eqref{eq:def-type-subst-7} }
  \end{align*}
  The remaining two cases are omitted.

@@

We have seen that a naive extension of substitution calculus with a universe breaks the local confluence of type substitution defined in \cref{subsec:SC-QIIRT}, because the newly added function clause interacts with other existing clauses.
In general, every term occurs on the right-hand side of a clause needs to propagate the reduction to allow a confluent term to exist, and this desired propagation can be achieved by turning a term involving a substitution $\sigma$ into a recursion.
We will see the same pattern in the next section.

@@

\begin{proposition} \label{prop:correctness-2}
  For every $\sigma : \Sub\;\Gamma\;\Delta$, and $A : \Ty\;\Delta$
  $\sigma \uparrow A$ is propositionally equal to $(\pi_1 \idS ; \sigma , \pi_2 \idS)$.
\end{proposition}
\begin{proof}
  We prove the statement by induction on $\sigma$ with \cref{lem:lifting}.
  Every case follows from equational reasoning straightforwardly.
  For example, if $\sigma = \sigma;\tau$ is a composite, we have
  \begin{align*}
    (\sigma;\tau) \uparrow A & \judgeq (\sigma \uparrow [\tau]\,A);(\tau \uparrow A) && \text{by definition} \\
                             & = \sigma^+;\tau^+                                     && \text{by induction hypothesis} \\ 
                             & = (\sigma;\tau)^+                                     && \text{by \cref{lem:lifting}.} 
  \end{align*}
  Other cases follow similarly.
\end{proof}
